---
title: "Extracting PRISM data for USACE Lakes"
author: "Iris Foxfoot"
date: "2024-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here) #package for managing working directories
library(janitor) #package for cleaning column names
library(sf) #package for managing spatial data
library(dplyr) #package for wrangling data
library(prism) #package for downloading prism climate data
library(terra) #package for handling rasters
```

# Purpose

In this rmarkdown I will download prism climate data, and extract values for each USACE reservoir.

# Data

- USACE_Reservoirs_Final.shp: A shape file containing all USACE reservoirs in the contiguous U.S.Downloaded from https://geospatial-usace.opendata.arcgis.com/datasets/03e322d7e89b48a9b48e9c3f4bcaf29e_0/explore. The data was modified in June 2024 to drop dry lakes, combine interconnected lakes, and drop small disconnected fragments.

The second dataset is PRISM 30 year (1990-2020) normal climate data, which was downloaded using the prism R package. For information on PRISM data see https://prism.oregonstate.edu/.

# Processing description

To produce this data, the center point of each reservoir was found. Using this center point, values for "ppt" (precipitation), "tmean" (mean temperature), "tmin" (min temperature), "tmax" (max temperature), "tdmean" (mean dew point temp) "vpdmin" (minimum vapor deficit), and "vpdmax" (max vapor deficit) were extracted from the PRISM 30 year (1990-2020) year normal 800m resolution raster cells. These data were then condensed into seasonal totals (for precipitation), seasonal means (for temperature, dew point--called dewp), seasonal minimums (for temperature and vapor pressure deficit --called vpd), and seasonal maximums (for temperature and vapor pressure deficit--called vpd)

# Cleaning reservoir data

```{r}
#load USACE reservoir data
res_raw <- read_sf("raw_data/USACE_Reservoirs/USACE_Reservoirs_Final.shp")


res_clean <- res_raw %>% 
  janitor::clean_names() %>% #convert column names to lower case snake
  sf::st_make_valid()

#get centers of lakes
res_center <- st_centroid(res_clean_name) %>% 
  select(objectid)

#plot data to make sure it looks okay
#plot(st_geometry(res_clean_name))
#plot(st_geometry(res_center), pch = 3, col = 'red', add = TRUE)
```

# Downloading PRISM data

```{r}
#set download directory
prism_set_dl_dir(here("raw_data/prism_data"))

#create function to download, process, and extract PRISM data
process_prism <- function(cli_var, lake_center){
  
  #if prism data doesn't exist for all 12 months, download it
  if(length(prism_archive_subset(cli_var, "monthly normals", resolution = "800m")) != 12) {
    get_prism_normals(type = cli_var, resolution = "800m",
                  mon = 1:12, annual = FALSE, keepZip = FALSE)
  }
  
  #get subset of prism data by type
  subset <- prism_archive_subset(cli_var, "monthly normals", resolution = "800m")
  
  #create raster stack
  stack <- pd_stack(subset)
  
  #make sure lake center is same projection as prism data
  lake_center_reproj <- st_transform(lake_center, st_crs(stack))
  
  #extract values for each lake center
  extracted <- raster::extract(stack, lake_center_reproj, df= T)
  
  #return extracted df
  return(extracted)
  
}

#create list of variables we want from prism normals dataset
climate_vars <- c(
  "ppt", #precipitation
  "tmean", #mean temperature
  "tmin", #min temperature
  "tmax", #max temperature
  "tdmean", #mean dew point temp
  "vpdmin", #minimum vapor deficit
  "vpdmax" #max vapor deficit
  )

#store new copy of res_clean for joining
res_with_prism <- res_clean

#loop through and join processed prism to lake data
#make sure you only run this once!!
for(i in climate_vars) {
  prism_dat <- process_prism(i, res_center) %>% select(-ID)
  res_with_prism <- cbind(as.data.frame(res_with_prism), prism_dat)
}
```

# Condense data into seasonal statistics 

```{r}
#condensing data by season
#winter = dec, jan, feb
#spring = mar, apr, may,
#summer = jun, jul, aug
#fall = sep, oct, nov

#function to condense data into seasonal summary
seasonal_summary <- function(x, season, data_type, fun) {
  season_months <- if(season == "winter") {c("12", "01", "02")} 
                   else if(season == "spring") {c("03", "04", "05")} 
                   else if(season == "summer") {c("06", "07", "08")}
                   else if(season == "fall") {c("09", "10", "11")}
  
  ls <- c()
  
  if(data_type == "ppt") {
    for (i in 1:3) {
      ls[i] <- paste0("PRISM_", data_type, "_30yr_normal_800mM4_", season_months[i], "_bil")
    }
  } else{
    for (i in 1:3) {
      ls[i] <- paste0("PRISM_", data_type, "_30yr_normal_800mM5_", season_months[i], "_bil")
    }
  }
    
  
  new_col <- 
    if (fun == "add") {
    x[,ls[[1]]] + x[,ls[[2]]] + x[,ls[[3]]]
      }
    else if (fun == "mean") {
    (x[,ls[[1]]] + x[,ls[[2]]] + x[,ls[[3]]])/3
    }
    else if (fun == "max") {
    pmax(x[,ls[[1]]], x[,ls[[2]]],x[,ls[[3]]])
    }
    else if (fun == "min")  {
    pmin(x[,ls[[1]]], x[,ls[[2]]],x[,ls[[3]]])
    }
  
  as.data.frame(new_col)

  return(new_col)
}


#get total precip per season
seasonal_climate_vals <- res_with_prism %>% 
  #seasonal total precip
  mutate(winter_total_precip = seasonal_summary(res_with_prism, season = "winter", data_type = "ppt", fun = "add")) %>% 
  mutate(spring_total_precip = seasonal_summary(res_with_prism, season = "spring", data_type = "ppt", fun = "add")) %>%
  mutate(summer_total_precip = seasonal_summary(res_with_prism, season = "summer", data_type = "ppt", fun = "add")) %>% 
  mutate(fall_total_precip = seasonal_summary(res_with_prism, season = "fall", data_type = "ppt", fun = "add")) %>% 
  #seasonal mean temp
  mutate(winter_mean_temp = seasonal_summary(res_with_prism, season = "winter", data_type = "tmean", fun = "mean")) %>% 
  mutate(spring_mean_temp = seasonal_summary(res_with_prism, season = "spring", data_type = "tmean", fun = "mean")) %>% 
  mutate(summer_mean_temp = seasonal_summary(res_with_prism, season = "summer", data_type = "tmean", fun = "mean")) %>% 
  mutate(fall_mean_temp = seasonal_summary(res_with_prism, season = "fall", data_type = "tmean", fun = "mean")) %>% 
  #seasonal min temp
   mutate(winter_min_temp = seasonal_summary(res_with_prism, season = "winter", data_type = "tmin", fun = "min")) %>% 
  mutate(spring_min_temp = seasonal_summary(res_with_prism, season = "spring", data_type = "tmin", fun = "min")) %>% 
  mutate(summer_min_temp = seasonal_summary(res_with_prism, season = "summer", data_type = "tmin", fun = "min")) %>% 
  mutate(fall_min_temp = seasonal_summary(res_with_prism, season = "fall", data_type = "tmin", fun = "min")) %>% 
  #seasonal max temp
     mutate(winter_max_temp = seasonal_summary(res_with_prism, season = "winter", data_type = "tmax", fun = "max")) %>% 
  mutate(spring_max_temp = seasonal_summary(res_with_prism, season = "spring", data_type = "tmax", fun = "max")) %>% 
  mutate(summer_max_temp = seasonal_summary(res_with_prism, season = "summer", data_type = "tmax", fun = "max")) %>% 
  mutate(fall_max_temp = seasonal_summary(res_with_prism, season = "fall", data_type = "tmax", fun = "max")) %>% 
  #seasonal mean dew point
  mutate(winter_mean_dewp = seasonal_summary(res_with_prism, season = "winter", data_type = "tdmean", fun = "mean")) %>% 
  mutate(spring_mean_dewp = seasonal_summary(res_with_prism, season = "spring", data_type = "tdmean", fun = "mean")) %>% 
  mutate(summer_mean_dewp = seasonal_summary(res_with_prism, season = "summer", data_type = "tdmean", fun = "mean")) %>% 
  mutate(fall_mean_dewp = seasonal_summary(res_with_prism, season = "fall", data_type = "tdmean", fun = "mean")) %>% 
 #seasonal min vapor deficit
  mutate(winter_min_vpd= seasonal_summary(res_with_prism, season = "winter", data_type = "vpdmin", fun = "min")) %>% 
  mutate(spring_min_vpd = seasonal_summary(res_with_prism, season = "spring", data_type = "vpdmin", fun = "min")) %>% 
  mutate(summer_min_vpd = seasonal_summary(res_with_prism, season = "summer", data_type = "vpdmin", fun = "min")) %>% 
  mutate(fall_min_vpd = seasonal_summary(res_with_prism, season = "fall", data_type = "vpdmin", fun = "min")) %>% 
   #seasonal max vapor deficit
  mutate(winter_max_vpd= seasonal_summary(res_with_prism, season = "winter", data_type = "vpdmax", fun = "max")) %>% 
  mutate(spring_max_vpd = seasonal_summary(res_with_prism, season = "spring", data_type = "vpdmax", fun = "max")) %>% 
  mutate(summer_max_vpd = seasonal_summary(res_with_prism, season = "summer", data_type = "vpdmax", fun = "max")) %>% 
  mutate(fall_max_vpd = seasonal_summary(res_with_prism, season = "fall", data_type = "vpdmax", fun = "max"))

seasonal_climate <- seasonal_climate_vals %>% 
  select(!PRISM_ppt_30yr_normal_800mM4_01_bil:PRISM_vpdmax_30yr_normal_800mM5_12_bil)
  

#save csv
st_write(seasonal_climate, here("extracted_data/usace_reservoir_seasonal_climate.csv"))

```


